---
title: "Data Wrangling in R"
author: "Yuan Hsiao"
date: "`r Sys.time()`"
output: beamer_presentation
---

# Acknowledgement: The slides have been modified from materials by 

* Chuck Lanfear (https://clanfear.github.io/CSSS508/)
* Stephanie Lee(http://students.washington.edu/syl3/csscr/)


---

# Data cleaning

For most statistical projects, the largest proportion of time spent is on **Data Cleaning**, which may involve:

* Generating variables that are relevant for analyses
* Subsetting the relevant observations that are relevant for analyses
* Getting the data in the right shape for the statistical program

---

# tidyverse and dplyr

We will be using the package **tidyverse** and its associated package **dplyr** in this workshop.

In the console: **install.packages("tidyverse")**

```{r}
library(tidyverse)
```

---

# Key functions in dplyr

* Four main functions
    * select() -- select columns
    * filter() -- select rows
    * mutate() -- add or transform variables
    * summarize() -- create summary statistics
    
* Two useful helper functions
    * `>%>` -- pipe operator: input into the next function
    * group_by() -- break into subsets

# Gapminder Data

We'll be working with data from Hans Rosling's [Gapminder](http://www.gapminder.org) project.
An excerpt of these data can be accessed through an R package called `gapminder`, cleaned and assembled by Jenny Bryan at UBC.

--

In the console: **install.packages("gapminder")**

Load the package and data:
```{r}
library(gapminder)
data("gapminder")
```

---

# A quick look at the data

```{r}
str(gapminder)
```

---

# What's Interesting Here?

* **Factor** variables `country` and `continent`

--

* Many observations: $n=`r nrow(gapminder)`$ rows

--

* A nested/hierarchical structure: `year` in `country` in `continent`

   + These are panel data!

---

# Pipes

`dplyr` allows us to use pipe operators (`%>%`) to "pipe" data between functions. So instead of nesting functions like this:

```{r}
log(mean(gapminder$pop))
```

--

We can pipe them like this:

```{r}
gapminder$pop %>% mean() %>% log()
```

--

Read this as, "send `gapminder$pop` to `mean()`, then send the output of that to `log()`."
In essence, pipes read "left to right" while nested functions read "inside to out."

---

# filter() -- Select rows that are relevant

Let's say we want observations from Oman after 1980 and through 2000.

--

```{r}
gapminder %>%
    filter(country == "Oman" &
           year > 1980 &
           year <= 2000 )
```

---

# Saving a Subset

If we think a particular subset will be used repeatedly, we can save it and give it a name like any other object:

```{r}
former_yugoslavia <- c("Bosnia and Herzegovina", 
                       "Croatia", 
                       "Macedonia", "Montenegro", 
                       "Serbia", "Slovenia")

yugoslavia <- gapminder %>% 
  filter(country %in% former_yugoslavia)

head(yugoslavia, 4)
```

---

# Sorting: arrange()

Along with filtering the data to see certain rows, we might want to sort it:

```{r}
yugoslavia %>% 
  arrange(year, desc(lifeExp))
```

The data are sorted by ascending **year** and descending **lifeExp**.

---

# Keeping Columns: select()

Not only can we limit rows, but we can include specific columns (and put them in the order listed) using `select()`. 

```{r}
yugoslavia %>% 
  select(country, year, pop) %>% 
    head(4)
```

---

# Dropping Columns: `select()`


We can instead drop only specific columns with **select()** using
**-** signs:

```{r}
yugoslavia %>% 
  select(-continent, -pop, -lifeExp) %>% 
    head(4)
```

---

# Create New Columns: `mutate()`

In `dplyr`, you can add new columns to a data frame using **mutate()**.

--


```{r}
yugoslavia %>% filter(country == "Serbia") %>%
    select(year, pop, lifeExp) %>%
    mutate(pop_million = pop / 1000000,
           life_exp_past_40 = lifeExp - 40) %>%
    head(5)
```

Note you can create multiple variables in a single `mutate()` call by separating the expressions with commas.

---

# Within `mutate()`: `case_when()`

**case_when()** allows you to create a new variable with values based on multiple logical statements. This is useful for making categorical variables or variables from combinations of other variables.

```{r}
gapminder %>% 
  select(gdpPercap) %>% 
    mutate(gdpPercap_ordinal = case_when(
      gdpPercap < 700 ~ "low",
      gdpPercap >= 700 & gdpPercap < 800 ~ "moderate",
      TRUE ~ "high" )) # Value when all other statements are FALSE
```

---

# Within `mutate()`: `recode()`

**recode()** is another useful function to use inside **mutate()**. Use **recode()** to change specific values to other values, particularly with factors. You can change multiple values at the same time. Note if a value has spaces in it, you'll need to put it in backticks!

```{r}
yugoslavia %>% 
  mutate(country = recode(country, 
                        `Bosnia and Herzegovina`="B and H", #<<
                        Montenegro="M")) %>% 
  distinct(country)
```

---

# summarizing in `dplyr`

Often we want to create summary statistics for the whole sample, or by groups (e.g., male, female). In these cases **summarize()** is a very useful function, often combined with **group_by()**

---

# summarize() Example

```{r}
yugoslavia %>%
    filter(year == 1982) %>%
    summarize(n_obs = n(),
              total_pop = sum(pop),
              mean_life_exp = mean(lifeExp),
              range_life_exp = max(lifeExp) - min(lifeExp))
```

These new variables are calculated using *all of the rows* in *yugoslavia*

---

# Grouping with `group_by()` 


The special function `group_by()` changes how functions operate on the data, most importantly `summarize()`.

Functions after `group_by()` are computed *within each group* as defined by variables given, rather than over all rows at once. Typically the variables you group by will be integers, factors, or characters, and not continuous real values.

---

# A `group_by()` example


```{r}
yugoslavia %>%
  group_by(year) %>% #<<
    summarize(num_countries = n_distinct(country),
              total_pop = sum(pop),
              total_gdp_per_cap = sum(pop*gdpPercap)/total_pop) %>%
    head(5)
```

Because we did `group_by()` with `year` then used `summarize()`, we get *one row per value of `year`*!

---

# Checkpoint: Small exercise

Task -- From the **gapminder** dataframe, we want the **average population in the millions** for each **country** in the continent **Europe**.

Step-by-step breakup -- From the **gapminder** dataframe:

* select relevant columns using **select()**
* select relevant rows using **filter()**
* create population in the millions using **mutate()**
* create average population in the millions for each country using **group_by()** with **summarize()**

---

```{r}
gapminder %>% 
  select(country, continent, pop) %>% 
    filter(continent == "Europe") %>% 
      mutate(pop_million = pop / 1000000) %>% 
        group_by(country) %>% 
          summarise(mean_pop_million = mean(pop_million))
```

---

# MERGING DATASETS -- `left_join()`

In many cases, we want to merge different sources of data into one dataframe.

* we can do this using **left_join()**

* In **left_join()**, we say rows should *match* because they have some columns containing the same value. 

* We list these in a **by = ** argument to the join.

---

# `nycflights13` Data

We'll use data in the [`nycflights13` package](https://cran.r-project.org/web/packages/nycflights13/nycflights13.pdf). 

Install and load it:
```{r}
# install.packages("nycflights13") # Uncomment to run
library(nycflights13)
```

It includes five dataframes, some of which contain missing data (`NA`):

* `flights`: flights leaving JFK, LGA, or EWR in 2013
* `airlines`: airline abbreviations
* `airports`: airport metadata
* etc

Note these are *separate data frames*, each needing to be *loaded separately*:

```{r, eval=FALSE}
data(flights)
data(planes)
data(airports)
# and so on...
```

---

# Join Example

Who manufactures the planes that flew to Seattle?

Take a look at the two dataframes we want to join: *flights* and *planes*

```{r}
head(flights)
```

---

```{r}
head(planes)
```

We want to join by the variable **tail_num**

---

```{r}
flights %>% 
  filter(dest == "SEA") %>% 
      left_join(planes %>% select(tailnum, manufacturer),
                by = "tailnum")
```

---

# Reshaping the data in the right format

In statistical analyses, we often want each row to be the unit of analysis (i.e., an observation). For example, take a look at the gapminder data

```{r}
head(gapminder)
```

Each row is a country-year, which is the basic unit of analysis

---

# Reshaping the data in the right format

What if the data looked like this?

```{r}
lifeExp_wide <- gapminder %>% 
  select(-pop, -gdpPercap) %>% 
  gather(variable, value, lifeExp) %>% 
    unite(var, variable, year) %>% 
      spread(var, value)

names(lifeExp_wide)
```

---

```{r}
head(lifeExp_wide)
```

---

# Reshaping the data in the right format -- **gather()**

* `gather()`: takes a set of columns and rotates them down to make two new columns (which you can name yourself): 
    * A **key** that stores the original column names
    * A **value** with the values in those original columns
    * Additional arguments to specify which columns you want
    
---

# `gather()` Example

```{r}
lifeExp_df1 <- lifeExp_wide %>% 
  gather(year, lifeExp, -country, -continent)

head(lifeExp_df1)
```

---

# `gather()` Example with `parse_number()`

```{r}
lifeExp_df1 <- lifeExp_wide %>% 
  gather(year, lifeExp, -country, -continent) %>% 
    mutate(year = parse_number(year))

head(lifeExp_df1)
```

---

# `spread()` does the opposite of `gather()`

For example:

```{r}
(too_long_data <- data.frame(Group = c(rep("A", 3), rep("B", 3)),
                             Statistic = rep(c("Mean", "Median", "SD"), 2),
                             Value = c(1.28, 1.0, 0.72, 2.81, 2, 1.33)))
```

---

# After `spread()`

```{r}
(just_right_data <- too_long_data %>%
    spread(key = Statistic, value = Value))
```

---

# We're done!

Questions?